---
title: "Lecture 02: Introduction to R"
author: "Timothy Chia"
date: "Tuesday May 12, 2026"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
---

```{r}
#| echo: false
options(repos = c(CRAN = "https://cloud.r-project.org"))
```

# Introduction

## Overview of R

R is a programming language developed in 1996 by Ross Ihaka and Robert Gentleman, two statistics professors from the University of Auckland in New Zealand.

R was initially intended to be merely a replacement for the statistics software used for teaching at their university department. However, R's evolution into a free, open-source language compatible with all major operating systems has led to its widespread adoption among statisticians and data scientists worldwide.

R's versatility is further augmented through a plethora of downloadable packages, containing functions and datasets that extend R's capabilities.

## Why is R so great?

R is a **functional programming language**, which means that functions are first-class objects that can be passed as arguments to other functions. However, R can support object-oriented programming, e.g., `ggplot2` is written using R objects.

Instead of covering how to write objects or functions in R, we will focus on how to use R's off-the-shelf functions and packages to perform data engineering and visualization.

## R Code in Quarto

R code can be integrated in a Quarto document using code chunks to generate documents combining text, code, and output.

The code chunk is defined as follows:

```{{r}}
#| eval: true
#| echo: true
#| output-location: column

# Your code here
```

Code execution options are specified using `#|` followed by the option name and value.

::: {.callout-note}
More information about Quarto code chunk execution options can be found [here](https://quarto.org/docs/output-formats/html-code-links.html).
:::

---

# Integrating Packages into R's Workflow

## Understanding Packages and CRAN

The base installation of R contains an extensive selection of built-in functions and datasets. Many R community members contribute additional functions and datasets beyond the base collection. Such additional collections of code are called **packages**.

R's official package repository is known as the **Comprehensive R Archive Network (CRAN)**. Currently, there are more than 20,000 packages available on CRAN.

## Managing the Package

In this course, we will work extensively with the `dplyr` package (included in `tidyverse`).

Print all loaded packages (compare this with `installed.packages()`):

```{r}
print(.packages())
```

If `dplyr` is not installed, install and load it using:

```{r}
install.packages("tidyverse") # `dplyr` is one of the `tidyverse` packages
library(dplyr)
```

Find the location of `dplyr`:

```{r}
find.package("dplyr")
```

`dplyr` contains a variety of functions and datasets, listed in the package documentation:

```{r}
help(package = "dplyr")
```

You may check the availability of `dplyr` by executing `starwars` in the Console tab:

```{r}
starwars
```

---

# Vectors

The fundamental data structure in R, known as a **vector**, encapsulates a combination of elements, all belonging to the same class, e.g., numbers or character strings.

## Creating Vector Variables (1-Based Indexing)

### Combining elements using `c()`

```{r}
c(31, -50, 93, 29, -44, 93)
```

### Shortcuts for numeric sequences

Create a numeric sequence using `:`:

```{r}
3:10
```

Create the index sequence of a given vector using `seq_along()`:

```{r}
seq_along(c(31, -50, 93, 29, -44, 93))
```

To create a sequence of a given length, use `seq_len()`:

```{r}
seq_len(5)
```

To create more general sequences, use `seq()`:

```{r}
seq(from = 5.2, to = -4, by = -0.5)
```

### The assignment operator

The assignment operator in R is `<-`:

```{r}
v <- c(31, -50, 93, 29, -44, 93)
v
```

### Vector length

Get the number of elements in a vector using `length()`:

```{r}
length(v)
```

Even a single element is treated as a vector:

```{r}
forty_two <- 42 # Do not write right-hand side as `c(42)`
length(forty_two)
```

## Classes of Vectors

### Numeric Vectors

Numeric vectors contain real-valued numbers. By default, vectors are numeric even if all elements are integers:

```{r}
v <- c(31, -50, 93, 29, -44, 93)
v
```

Get the class of a vector using `class()`:

```{r}
class(v)
```

Check if `v` is numeric using `is.numeric()`:

```{r}
is.numeric(v)
```

### Integer Vectors

Check if `v` is integer:

```{r}
is.integer(v)
```

Convert `v` to integer using `as.integer()`:

```{r}
w <- as.integer(v)
w
```

Check if `w` is integer:

```{r}
is.integer(w)
is.numeric(w)
```

### Character Vectors

```{r}
compass <- c(
  "North", "North East", "East",
  "South East", "South",
  "South West", "West",
  "North West"
)
compass
```

Check the class of `compass`:

```{r}
class(compass)
```

**Special built-in character vectors:**

```{r}
LETTERS
```

```{r}
letters
```

### Logical Vectors

Logical vectors contain `TRUE` or `FALSE` values:

```{r}
agree <- c(TRUE, TRUE, FALSE, FALSE, TRUE)
class(agree)
class(TRUE)
class("TRUE")
```

No quotation marks:

```{r}
class(agree)
is.logical(agree)
```

## Class Casting (Coercing)

> **Question:** Can we mix different classes in a single vector?

To cast an object to a different class, use:

- `as.numeric()`
- `as.integer()`
- `as.character()`
- `as.logical()`
- `as.factor()`

::: {.callout-note}
More information about coercion can be found [here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/coerce.html).
:::

## Accessing Vector Elements

### Extracting Vector Elements

```{r}
compass
```

```{r}
compass[1:2]
```

```{r}
compass[c(2, 8, 4, 4)]
```

```{r}
head(compass, 2)
```

```{r}
tail(agree, 3)
```

### Positive and Negative Indexing

```{r}
compass[c(2, 4, 6, 8)]
```

```{r}
compass[-c(2, 4, 6, 8)]
```

### Logical Indexing

```{r}
agree[c(TRUE, FALSE, TRUE, FALSE, TRUE)]
```

```{r}
agree[c(FALSE, TRUE)]
```

### Replacing Vector Elements

```{r}
compass_code <- compass
compass_code[1] <- "N"
compass_code
```

```{r}
compass_code[c(5, 8)] <- c("S", "NW")
compass_code
```

```{r}
compass_code[c(2:4, 6:7)] <- c("NE", "E", "SE", "SW", "W")
compass_code
```

---

# Passing Arguments to Functions

Let us now explore different methods for passing arguments to R functions.

## Arguments with Default Values

The `pull()` function in `dplyr` extracts a single column from a data frame and returns it as a vector. The 2nd and 3rd argument (`var` and `name`) are followed by `=`. The absence of `=` for `.data` implies that it does not have a default value.

Using the default argument values:

```{r}
head(pull(starwars, height))
```

It is equivalent to:

```{r}
head(starwars$height)
```

## Argument Matching by Name

Specify the argument values:

```{r}
head(pull(.data = starwars, var = eye_color, name = name), 5)
```

Default values do not need to be explicitly specified:

```{r}
p_1 <- pull(.data = starwars, var = eye_color, name = NULL)
p_2 <- pull(.data = starwars, var = eye_color)
identical(p_1, p_2)
```

Arguments without default values must be specified:

```{r}
# pull(var = eye_color, name = NULL) # Error: object 'eye_color' not found
```

The order of the arguments does not matter when using names:

```{r}
p_3 <- pull(.data = starwars, var = eye_color, name = name)
p_4 <- pull(name = name, .data = starwars, var = eye_color)
identical(p_3, p_4)
```

## Argument Matching by Position

When matching by position, the arguments can be omitted:

```{r}
p_5 <- pull(.data = starwars, var = eye_color, name = name)
p_6 <- pull(starwars, eye_color, name)
identical(p_5, p_6)
```

When using mixed matching, once an argument is matched by name, the remaining arguments should also be matched by name. Otherwise, the code would be difficult to read.

```{r}
head(pull(starwars, eye_color, name = name), 5)
```

**tidyverse style guide stylistically preferred version:**

```{r}
head(pull(starwars, var = eye_color, name = name))
```

## Pipes (Important!)

Data analysis often involves a lengthy sequence of function calls. Consider the following sequence of steps:

1. Take the built-in vector `rivers`, which contains the lengths of 141 major North American rivers in miles.
2. Extract a random subset of five values from `rivers` using the `sample()` function.
3. Calculate the sum of the subset.
4. Compute the cumulative distribution function for the normal distribution (mean 2000, sd 1000), evaluated at the sum.
5. Print the result, showing five digits after the decimal point.

This sequence can be implemented using the following code:

```{r}
tmp_1 <- sample(rivers, 5)
tmp_2 <- sum(tmp_1)
tmp_3 <- pnorm(tmp_2, mean = 2000, sd = 1000)
print(tmp_3, digits = 5)
```

**Disadvantage:** `tmp_1`, `tmp_2`, and `tmp_3` are required for temporary storage. Possible to eliminate them a long chain of function calls:

```{r}
print(pnorm(sum(sample(rivers, 5)), mean = 2000, sd = 1000), digits = 5)
```

**Disadvantage:** Hard to read and analyze. R's pipe operator `|>` was introduced in 2021, which means "then" in the code flow.

**Pipe version:**

```{r}
rivers |>
  sample(5) |>
  sum() |>
  pnorm(mean = 2000, sd = 1000) |>
  print(digits = 5)
```

:::{.callout-note}
Avoid using pipes for simple operations, e.g., a single function call. When there are too many pipes, consider creating intermediate variables.
:::


---

# Numeric Operations

Numeric operations in R are the basic arithmetic operations that can be performed on vectors and other data structures.

## Numeric Operations

R's numeric operations work on vectors element-wise by default. When two vectors of the same length are operated on, R applies the operation to each corresponding pair of elements.

When vectors are of different lengths, R recycles (repeats) the shorter vector until it matches the length of the longer one. If the longer length is not a multiple of the shorter length, R issues a warning.

### Element-wise operations

R applies arithmetic operators to vectors **element by element**. Given two vectors of the same length, the operation is performed on each corresponding pair of elements.

```{r}
x <- c(9, 5, 4, 5)
y <- c(4, 9, 3, 6)
x + y
```

```{r}
x / y
```

### Mathematical functions

**Broadcasting a scalar:**

When a scalar is combined with a vector, R applies the operation to **every element** of the vector. This is known as broadcasting.

```{r}
x + 3
```

**Broadcasting a vector:**

When a vector is combined with another vector of the same length, R applies the operation element-wise.

```{r}
x + c(1, -1, -1)
```

**Line should break after an operator:**

R allows long expressions to be split across multiple lines. The convention is to break **after** an operator so the continuation is visually clear.

```{r}
1 + 2 + 3 + 4 + 5 + 6 + 7 + 8 + 9 + 10 +
  11 + 12 + 13 + 14 + 15 + 16 + 17 + 18 + 19 + 20
```

**Mathematical function examples:**

R provides built-in mathematical functions that accept **vector inputs** and return vector outputs:

```{r}
sqrt(c(9, 16, 25))
exp(0:3)
log10(c(1, 10, 100))
```

### Broadcasting

**Vector recycling** occurs when two vectors of different lengths are combined. R repeats (recycles) the shorter vector until it matches the length of the longer one.

:::{.callout-warning}
If the longer length is not an exact multiple of the shorter length, R issues a warning.
:::

```{r}
r <- seq(from = 6, to = 2, by = -2)
s <- 3:1
r[r - s]
```

```{r}
t <- c(TRUE, TRUE, FALSE, FALSE, FALSE)
u <- c(FALSE, FALSE, TRUE, TRUE, FALSE)
t[u]
```

```{r}
t <- c(TRUE, TRUE, FALSE, FALSE, FALSE)
u <- c(TRUE, FALSE, TRUE)
t[u]
```
::: {.callout-note}
When the shorter vector is recycled, its elements are repeated from the beginning.
:::

```{r}
t <- c(TRUE, TRUE, FALSE, FALSE, FALSE)
u <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
t[u]
```

## Appendix — List of Numeric Operations

This appendix provides a quick reference for the most commonly used numeric operations in R.

### Basic Arithmetic Operations

R supports all standard arithmetic operators. Most of these work element-wise on vectors, applying the operation to each pair of corresponding elements. The modulo operator `%%` and integer division operator `%/%` are particularly useful for data manipulation tasks such as checking divisibility or extracting specific digits.

| Operator | Description | Example Call | Example Result |
|----------|-------------|--------------|----------------|
| `x + y` | x plus y | `3 + 5` | `8` |
| `x - y` | x minus y | `10 - 4` | `6` |
| `x * y` | x multiplied by y | `3 * 7` | `21` |
| `x / y` | x divided by y | `10 / 4` | `2.5` |
| `x^y` or `x**y` | x raised to the power y | `2^3` | `8` |
| `x %/% y` | Integer division | `7 %/% 3` | `2` |
| `x %% y` | x modulo y | `7 %% 3` | `1` |

### Basic Summary Statistics

R provides a comprehensive set of built-in functions for computing summary statistics. These functions accept vector inputs and return a single numeric value that describes some aspect of the data. Most of these functions have an `na.rm` argument that can be set to `TRUE` to ignore missing values.

| Function | Description | Example Call | Example Result |
|----------|-------------|--------------|----------------|
| `length()` | Number of elements | `length(c(1, 0, 8, 2, 1))` | `5` |
| `sum()` | Sum | `sum(c(1, 0, 8, 2, 1))` | `12` |
| `mean()` | Mean (average) | `mean(c(1, 0, 8, 2, 1))` | `2.4` |
| `median()` | Median | `median(c(1, 0, 8, 2, 1))` | `1` |
| `sd()` | Standard Deviation | `sd(c(1, 0, 8, 2, 1))` | `3.209` |
| `var()` | Variance | `var(c(1, 0, 8, 2, 1))` | `10.3` |
| `IQR()` | Interquartile range | `IQR(c(1, 0, 8, 2, 1))` | `1` |
| `min()` | Minimum | `min(c(1, 0, 8, 2, 1))` | `0` |
| `max()` | Maximum | `max(c(1, 0, 8, 2, 1))` | `8` |
| `range()` | Range | `range(c(1, 0, 8, 2, 1))` | `c(0, 8)` |
| `quantile()` | Quantiles | `quantile(c(1, 0, 8, 2, 1), probs = c(0.25, 0.75))` | `25% = 1, 75% = 2` |
| `cov()` | Covariance | `cov(c(1, 0, 8, 2, 1), 1:5)` | `0.5` |
| `cor()` | Correlation | `cor(c(1, 0, 8, 2, 1), 1:5)` | `0.099` |

### Basic Mathematical Functions

R includes a wide range of mathematical functions that operate on vectors element-wise. These include absolute value, roots, factorials, exponentials, logarithms, and trigonometric functions. R also provides functions for working with the normal distribution, which are essential for statistical analysis and hypothesis testing.

| Function | Description |
|----------|-------------|
| `abs(x)` | Absolute value |
| `sqrt(x)` | Square root |
| `factorial(x)` | Factorial |
| `exp(x)` | Exponential function |
| `log(x)` | Natural logarithm |
| `log10(x)` | Logarithm with base 10 |
| `cos(x)`, `sin(x)`, `tan(x)` | Trigonometric functions |
| `acos(x)`, `asin(x)`, `atan(x)` | Inverse trigonometric functions |
| `dnorm(x, mean, sd)` | Density function of the normal distribution |
| `pnorm(x, mean, sd)` | Cumulative distribution function of the normal distribution |

---

# Logical and Comparison Operators

R provides a variety of logical and comparison operators that can be used to perform operations on vectors and other data structures.

## Logical Operators 

Logical operators in R work **element-wise** on logical vectors. The three primary logical operators are:

- `&` — **AND**: returns `TRUE` only when both corresponding elements are `TRUE`
- `|` — **OR**: returns `TRUE` when at least one of the corresponding elements is `TRUE`
- `!` — **NOT**: inverts the logical value of each element

> Example: Analyzing website login attempts using two-factor authentication (2FA).

```{r}
password <- c(TRUE, TRUE, FALSE, FALSE)
two_factor <- c(FALSE, TRUE, FALSE, TRUE)
```

Access is granted only if both the password and 2FA are correct:

```{r}
password & two_factor
```

If access is granted given at least one correct input:

```{r}
password | two_factor
```

NOT operator: `!password` indicates whether a customer used a wrong password:

```{r}
!password
```

**Combining logical conditions:**

Logical operators can be combined to form complex conditions. For example, to find customers who are female, younger than 40, and neither married nor divorced:

```r
"female" & age < 40 & !(status %in% c("married", "divorced"))
```

> Question: What is the difference between `&` and `&&`?

## Comparison Operators

Comparison operators compare two vectors element-wise and return a logical vector. The operators are:

| Operator | Description: TRUE if and only if … |
|----------|-------------------------------------|
| `x == y` | x equals y |
| `x != y` | x does not equal y |
| `x < y` | x is less than y |
| `x > y` | x is greater than y |
| `x <= y` | x is less than or equal to y |
| `x >= y` | x is greater than or equal to y |
| `dplyr::between(x, left, right)` | x is between left and right, inclusively |
| `x %in% y` | x is an element in y |

**Example — comparing vectors of different lengths:**

```{r}
vector1 <- c(10, 20, 30, 40)
vector2 <- c(15, 20, 25, 40, 30)
vector3 <- c(10, 20, 30, 40, 30)
vector1 == vector2
```
::: {.callout-warning}
Be careful when working with vectors of different lengths. R recycles the shorter vector.
:::

**Checking if two vectors are exactly the same:**

```{r}
identical(vector1, vector3)
all(vector1 == vector2)
sum(vector1 - vector2) == 0
```

**Numbers divisible by 2:**

```{r}
numbers <- c(31, -50, 93, 29, 12, 8)
numbers %% 2 == 0
```

**String comparison:**

```{r}
"farewell" == "welcome"
"farewell" < "welcome"
```
::: {.callout-note}
Alphabetical order depends on the user's "locale," a set of parameters from which R infers the user's location. In Danish, "z" falls between "y" and "aa", whereas in Estonian "z" comes between "s" and "t". Throughout this course, we adhere to standard English alphabetical order.
:::

---

# Missing Values

Missing values are common in data workflow. They can be caused by missing information during data collection, entry errors, or other reasons. In R, missing values can be represented by `NA`, `NaN`, or `NULL`.

## `NA`, `NaN`, and `NULL`

| Value | Meaning |
|-------|---------|
| `NA` | Exists in principle but not available, e.g., missing ages in titanic |
| `NaN` | Exists in principle but not a number, e.g., `sqrt(-3)` |
| `NULL` | Non-existent object |

`NA` can adjust its class to match the context:

```{r}
x <- c(1, 2, NA, 5)
is.na(x)
```

`NaN` (Not a Number) results from undefined mathematical operations:

```{r}
0 / 0
```

```{r}
Inf / Inf
```

```{r}
log(-1)
```

`NULL` represents a non-existent object. Unlike `NA`, `NULL` has length 0:

```{r}
z <- c(1, Inf, NULL, NaN, NA)
z
is.nan(z)
```

```{r}
a <- NULL
exists("a")
is.null(a)
```
::: {.callout-note}
Never use `== NULL`. Use `is.null()` instead.

Comparing with `== NULL` does not work as expected in R. For example, when doing `x == NULL`, R recycles `NULL` and returns `logical(0)` (an empty logical vector) rather than `FALSE`. This can lead to subtle bugs in conditional statements where you expect a single `TRUE` or `FALSE` value.
:::

**Handling NA in arithmetic:**

When a vector contains `NA`, most operations return `NA` unless explicitly told to ignore missing values:

```{r}
mean(x)
mean(x, na.rm = TRUE)
```

Check for any missing values using `anyNA()`:

```{r}
anyNA(x)
```

## Handling Missing Values: In-Class Exercise

Run the following code on your computer and calculate the mean values of `df1` and `df2`, respectively. As a team, try to find the right built-in function for calculation and the right argument to deal with missing values. What do you observe from the results?

```{r}
set.seed(0)

df <- rnorm(100, mean = 5, sd = 5) |>
  format(digits = 2) |>
  as.numeric()

index1 <- sample(1:100, 3)
index2 <- sample(1:100, 3)

df1 <- df
df2 <- df

df1[index1] <- NA
df2[index2] <- NaN
```

---

# Conditional Element Selection

R provides several functions for conditional element selection, including `dplyr::if_else()`, `dplyr::case_when()`, and `dplyr::case_match()`.

## `if_else()`

`if_else()` is a **vectorised** version of the base R `ifelse()`. It takes a logical condition and returns one value if `TRUE` and another if `FALSE`, applied element-wise.

```{r}
v1 <- c(31, -50, 93, NA, 29, -44, 93)
v1 %% 2 == 0
```

```{r}
if_else(v1 %% 2 == 0, "Yes", "No")
```

`if_else()` also works with numeric outputs:

```{r}
if_else(v1 %% 2 == 0, seq_along(v1), 0)
```

**Nested `if_else()`** can handle multiple conditions, but becomes hard to read:

```{r}
v2 <- c(31, -50, 93, NA, 29, -44, 93)
if_else(
  v2 %% 2 == 0, "even",
  if_else(v2 %% 2 == 1, "odd", "neither even nor odd")
)
```

## `case_when()`

For multiple conditions, `case_when()` provides a cleaner, more readable syntax using the `~` (tilde) operator:

```{r}
case_when(
  v2 %% 2 == 0 ~ "even",
  v2 %% 2 == 1 ~ "odd",
  !is.na(v2) ~ "neither even nor odd"
)
```

## `case_match()`

`case_match()` is useful when matching exact values rather than conditions:

```{r}
directions <- c(
  "east", "west", "north", "north", NA, "east",
  "south", "west", "south", "east", "north", "zigzag"
)
```

---

**Disadvantage:** code repetition with nested `if_else()`:

```{r}
#| eval: false
if_else(
  directions == "north" ~ "N",
  directions == "east" ~ "E",
  directions == "south" ~ "S",
  directions == "west" ~ "W"
)
```

---

**Recommendation:** Use `if_else()` in single-condition and `case_when()` and `case_match()` in multiple conditions. An alternative to `dplyr`'s `if_else()` is the built-in function `ifelse()`. However, `if_else()` offers faster and more intuitive results.

```{r}
case_when(
  directions == "north" ~ "N",
  directions == "east" ~ "E",
  directions == "south" ~ "S",
  directions == "west" ~ "W"
)
```

```{r}
case_match(
  directions,
  "north" ~ "N",
  "east" ~ "E",
  "south" ~ "S",
  "west" ~ "W"
)
```

---

# Writing Functions

We are not going to cover all the details of writing functions in R. However, we will introduce some basic concepts and provide some examples, and you should be aware of how an R function is structured.

## Function Basics

```{r}
library(stringr)
```

```{r}
say_hello <- function(name) {
  print(str_c("Hello, ", name, "!"))
}
say_hello("Parallel Universe")
```

## Default Values

Functions can have default parameter values:

```{r}
say_hello <- function(name = "World") {
  print(str_c("Hello, ", name, "!"))
}
say_hello()
```

## `return()` Function

**Explicit return:**

```{r}
increment <- function(x) {
  return(x + 1)
}
increment(5)
```

**Optional `return()`:**

R implicitly returns the last evaluated expression, so `return()` is optional:

```{r}
increment <- function(x) {
  x + 1
}
increment(5)
```

**Return `NULL` example:**

Some functions return `NULL` implicitly (e.g., `message()`, `print()`):

```{r}
my_return_null_example <- function() {
  message("I am showing a lot of")
}
z <- my_return_null_example()
z
```

::: {.callout-note}
Check tidyverse style guide for recommended practices of using `return()`.
:::

# Data Frames (tibble)

## Data Frames (tibble) (1) — Creating Data Frames

Data Frame: rectangular shaped table. Columns: variables. Rows: observations. Simply put, a data frame is a well-formatted table.

### Row-Wise Creation

`tibble::tribble()` allows you to create data frames by specifying rows one at a time. Column names are preceded by `~`:

```{r}
timetable <- tibble::tribble(
  ~Week, ~`Monday (9–11am)`, ~`Thursday (9–11am)`,
  1, "Intro to `RStudio` and `Quarto`", "Software Setup",
  2, "Intro to `R`", "**Assessment 1**",
  3, "`tidyverse` Single-Table Queries", "Team Project Description",
  4, "`tidyverse` Join-Table Queries", "**Assessment 2**",
  5, "Time Series Basics", "Team Project Q&A",
  6, "Data Engineering Case Study", "**Assessment 3**",
  8, "Data Visualization Using `ggplot2`", "Team Project Q&A",
  9, "Exploring `geom`", "**Assessment 4**",
  10, "Aesthetic Mappings", "Team Project Q&A",
  11, "Intro to `gt` Tables", "**Assessment 5**",
  12, "Visualizing Geospatial Data", "Team Project Q&A",
  13, "Team Project Evaluation", "Team Project Evaluation"
)
timetable
```

### Column-Wise Creation

`tibble()` creates data frames column by column, similar to `data.frame()`:

```{r}
ny_mayor <- tibble(
  borough = c("Bronx", "Brooklyn", "Manhattan", "Queens", "Staten Island"),
  de_blasio = c(117712, 254755, 190312, 171867, 25466),
  malliotakis = c(23715, 74343, 53853, 94911, 70125),
  other = c(3138, 10602, 10186, 8974, 2747)
)
ny_mayor
```

Both approaches produce tibbles with the same class:

```{r}
class(timetable)
class(ny_mayor)
```

## Data Frames (tibble) (2) — Working with Data Frames

### Extracting Data Frame Content

There are multiple ways to extract content from a data frame:

**Using the `$` operator:**

```{r}
ny_mayor$other
```

```{r}
ny_mayor$"other"
```

**Using double brackets `[[ ]]` (returns a vector):**

```{r}
ny_mayor[["other"]]
```

```{r}
ny_mayor[[4]]
```

**Using single brackets `[ ]` (returns a data frame):**

```{r}
ny_mayor[4]
```

```{r}
ny_mayor["other"]
```

::: {.callout-note}
`ny_mayor[other]` will throw an error (`object 'other' not found`) because R interprets `other` as a variable name, not a column name.
:::

**Data frame metadata:**

```{r}
nrow(timetable)
ncol(timetable)
dim(timetable)
colnames(timetable)
```

```{r}
summary(ny_mayor)
```

### Slice and Filter

**Viewing rows:**

```{r}
slice_head(timetable, n = 2)
```

```{r}
slice_tail(timetable, n = 2)
```

::: {.callout-tip}
Try `str(timetable)` and `glimpse(timetable)` to inspect the structure of a data frame.
:::

### `pull()`, `select()`, `slice()`, `filter()`

**`pull()` — extract a single column as a vector:**

```{r}
pull(ny_mayor, var = de_blasio, name = borough)
```

**Subsetting with base R:**

```{r}
ny_mayor[3:5, "de_blasio"]
```

```{r}
ny_mayor[[3, "de_blasio"]]
```

::: {.callout-warning}
`ny_mayor[[3:5, "de_blasio"]]` will throw an error — `[[` expects a single row index, not a vector.
:::

**Filtering with `filter()`:**

```{r}
de_blasio_strongholds_1 <-
  ny_mayor[ny_mayor$de_blasio > ny_mayor$malliotakis, ]
de_blasio_strongholds_2 <-
  filter(ny_mayor, de_blasio > malliotakis)
```

```{r}
identical(de_blasio_strongholds_1, de_blasio_strongholds_2)
```

## Data Frames (tibble) (3) — Subsetting

Subsetting data frames can be done with base R or tidyverse functions. All approaches below produce the same result:

**Column subsetting:**

```{r}
ny_dfr_1 <- ny_mayor[1:2]
ny_dfr_2 <- ny_mayor[c("borough", "de_blasio")]
ny_dfr_3 <- select(ny_mayor, 1:2)
ny_dfr_4 <- select(ny_mayor, borough, de_blasio)
```

```{r}
identical(ny_dfr_1, ny_dfr_2)
identical(ny_dfr_1, ny_dfr_3)
identical(ny_dfr_1, ny_dfr_4)
```

**Row subsetting:**

```{r}
first_three_rows_1 <- ny_mayor[2:3, ]
first_three_rows_2 <- slice(ny_mayor, 2:3)
identical(first_three_rows_1, first_three_rows_2)
```

## Data Frames (tibble) (4) — Modifying Data Frames

### Appending Columns

Use `mutate()` to add new columns based on existing ones:

```{r}
mutate(
  ny_mayor,
  total = de_blasio + malliotakis + other,
  de_blasio_pct = 100 * de_blasio / total
)
```

### Replacing Columns

`mutate()` can also replace existing columns:

```{r}
mutate(
  ny_mayor,
  de_blasio = de_blasio * 1000,
  malliotakis = malliotakis + 1
)
```

### Appending Rows

Use `add_row()` to append new rows:

```{r}
fictitious_ny <- add_row(
  ny_mayor,
  tribble(
    ~borough, ~de_blasio, ~malliotakis, ~other,
    "Bagelburg", 1, NA, 4,
    "Hipster Haven", 2, 3, 5
  )
)
slice_tail(fictitious_ny, n = 4)
```

### Replacing Rows

Use `rows_update()` to replace rows by a key column:

```{r}
rows_update(
  ny_mayor,
  tribble(
    ~borough, ~de_blasio, ~malliotakis, ~other,
    "Manhattan", 0, 1, 2,
    "Staten Island", 3, NA, 4
  ),
  by = "borough"
) |>
  slice(2:5,)
```

## Data Frames (tibble) (5) — Advanced Manipulation

### Replacing Arbitrary Cells

Base R indexing can replace individual cells or blocks of cells:

```{r}
ny_mayor
```

```{r}
ny_mayor_copy <- ny_mayor
ny_mayor_copy[1, "de_blasio"] <- 0
```

```{r}
ny_mayor_copy[3:4, c(1, 3)] <- tribble(
  ~borough, ~malliotakis,
  "Bagelburg", 1,
  "Hipster Haven", NA
)
ny_mayor_copy
```

### Renaming Columns

**Using `rename()`:**

```{r}
rename(ny_mayor, democrat = de_blasio, republican = malliotakis)
```

**Using `select()` with renaming syntax:**

```{r}
select(ny_mayor, democrat = de_blasio, republican = malliotakis, other)
```

---

# Factors

## Factors (1) — Types of Data

`factor` is R's data structure for representing categorical data. There are three main types of data that factors can represent:

1. **Identifiers:** Unique values associated with individual observations. Examples include:
   - Names of US states
   - Passport numbers
   - Bayer names that identify stars in the sky

2. **Quantitative data:** Numbers that allow meaningful arithmetic operations. Example:
   - US state area (available in R as built-in numerical `state.area` vector)
   - Height of a person
   - Diameter of a star

3. **Categorical data:** Observations split into a small number of groups. Example:
   - Region of a US state (Northeast, South, North Central, West)
   - Undergraduate years in college (Year 1, Year 2, Year 3, Year 4)
   - The grades you can have (A, B, C, D, F)

## Factors (2) — Working with Factors

```{r}
library(forcats)
```

**Creating factors:**

```{r}
factor(c("Pomposo", "Colossale", "Pomposo", "Regolare"))
```

**Setting custom levels:**

```{r}
sizes <- factor(
  c("Pomposo", "Colossale", "Pomposo", "Regolare"),
  levels = c("Modesto", "Regolare", "Pomposo", "Colossale")
)
sizes
```

```{r}
sizes_with_piccolo <- factor(
  c("Colossale", "Regolare", "Modesto", "Piccolo"),
  levels = c("Modesto", "Regolare", "Pomposo", "Colossale")
)
sizes_with_piccolo
```

```{r}
levels(sizes)
nlevels(sizes)
```

**Recode factor levels with `forcats::fct_recode()`:**

```{r}
abbreviated_sizes <- fct_recode(
  sizes,
  m = "Modesto",
  r = "Regolare",
  p = "Pomposo",
  c = "Colossale"
)
abbreviated_sizes
```

**Collapse same category with different names using `forcats::fct_collapse()`:**

```{r}
sizes_with_big <- factor(c("Pomposo", "Big", "r"))
fct_collapse(sizes_with_big, p = c("Pomposo", "Big"))
```

## Factors (3) — Use Cases

### Counting and Sorting Levels by Frequency

```{r}
str(gss_cat)
```

```{r}
levels(gss_cat$race)
```

```{r}
fct_count(gss_cat$race)
```

```{r}
fct_count(fct_infreq(gss_cat$race))
```

::: {.callout-note}
`gss_cat` contains 21,483 responses in the General Social Survey. This biannual survey, conducted by the National Opinion Research Center at the University of Chicago, collects demographic information from representative respondents in the US population, along with their opinions, attitudes, and behaviors.
:::

## Factors (4) — Advanced Factor Operations

### Drop Unused Factor Levels

```{r}
races <- gss_cat$race
levels(fct_drop(races))
```

### Ordered Factors

Nominal factors cannot be compared:

```{r}
sizes > "Regolare"
```

Ordered factors can be compared:

```{r}
sizes_ordered <- factor(
  c("Pomposo", "Colossale", "Pomposo", "Regolare"),
  levels = c("Modesto", "Regolare", "Pomposo", "Colossale"),
  ordered = TRUE
)
sizes_ordered > "Regolare"
```

---

# Importing Data — CSV and XLSX

Download the `epl.zip` file from the course website (Content → Data), to your work directory. It contains a `.csv` and a `.xlsx` file of the same data, respectively.

The files can be imported into RStudio by either using the GUI or running R commands.

**GUI:** Click the file name and select **Import Dataset…**

**R command:**

```{r}
# readr::read_csv() for CSV files
# readxl::read_xlsx() for XLSX files
# readxl::read_excel() guesses whether the file is XLS or XLSX
```

::: {.callout-note}
`readxl::read_excel()` guesses whether the file is XLS or XLSX. More information [here](https://readxl.tidyverse.org/). While `readxl::read_xls()` or `readxl::read_xlsx()` work with known extensions.
:::

---

# Code Styling and Obtaining Help

We introduce two packages for code styling: `styler` and `lintr`. We also provide some tips on obtaining help when you encounter problems in R.

## Styling Code Using the `styler` Package

A consistent code style is a sign of a professional attitude to programming.

The `styler` package installs an RStudio menu option that automatically edits R code so that it adheres to the tidyverse style. It is not a perfect substitute for reading the tidyverse style guide. However, it can implement many aspects of the tidyverse style, such as proper indentation and using the assignment operator `<-` instead of an equals sign `=` for assignments.

Install `styler` and use the **Style active file** option from the **Addins** menu to style your code:

```{r}
install.packages("styler")
```

## Linting Code Using the `lintr` Package

`lintr` is a static code analysis tool that checks R code for common style issues and possible semantic problems.

Like `styler`, `lintr` is not a substitute for reading the tidyverse style guide, but it can help to identify potential problems in your code.

**Installation and usage of `lintr`:**

```{r}
#| eval: false
install.packages("lintr")
lintr::lint("week_02/week_02_intro_to_r.qmd")
```

If any problems were detected in the file, they will be displayed in the **Markers** tab in the bottom left RStudio pane. Clicking any of the detected problems will take you to the corresponding line in the source code.

::: {.callout-note}
It is advisable to fix as many issues as possible before sharing your code with others. Under rare circumstances, you may want to suppress a warning by adding the comment `# nolint` to the end of the corresponding line of code.
:::

```r
x <- 3 # nolint
```

---

# Summary

We have covered a series of essential R skills:

- **R** is a versatile programming language that is widely used in statistics and data science.
- The **fundamental data structure** in R is the **vector**, which can contain elements of the same class.
- **Arguments** can be passed to functions by name, position, or both.
- **Pipe operator `|>`** can be used to chain together a sequence of function calls in R.
- **Vectors** can be created using the `c()` function or by using shortcuts such as `:` and `seq()`.
- **Operators**: Numerical, logical (`&`, `|`, and `!`), and comparison.
- **Missing Values**: `NA`, `NaN`, and `NULL`.
- **Conditional Element Selection**: `if_else()`, `case_when()`, and `case_match()`.
- **Writing Functions**: `function()` and `return()`.
- **Data Frames (tibble)**: creation, manipulation, and extraction.
- **Factors**: categorical data representation.
- **Importing Data**: CSV and XLSX files.
- The **`styler`** and **`lintr`** packages can be used to automatically format R code according to the tidyverse style guide.
